---
title: "Exercise"
author: "ktrebach"
date: "January 24, 2016"
output: html_document
      keep_md: true
---

Review criteria
What you should submit

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

Peer Review Portion

Your submission for the Peer Review portion should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).

Course Project Prediction Quiz Portion

Apply your machine learning algorithm to the 20 test cases available in the test data above and submit your predictions in appropriate format to the Course Project Prediction Quiz for automated grading.

Reproducibility

Due to security concerns with the exchange of R code, your code will not be run during the evaluation by your classmates. Please be sure that if they download the repo, they will be able to view the compiled HTML version of your analysis.

Prediction Assignment Writeupless 
Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

Data

The training data for this project are available here:
      
      https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:
      
      https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv


The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

```{r loaddata, echo=FALSE, message=FALSE, cache=TRUE}

urltrain <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
urltest <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
setwd("~/Desktop/Data Scientist JH MOOC/Practical Machine/gitrepos4/practicalmachinelearning")

missing.types <- c("NA", "") # assign NA to all NA and all "".


download.file(urltrain,destfile="./pmltraining.csv",method="curl")
training <- read.csv("./pmltraining.csv",stringsAsFactors=TRUE, na.strings=missing.types)

download.file(urltest,destfile="./pmltest.csv",method="curl")
test <- read.csv("./pmltest.csv",stringsAsFactors=TRUE, na.strings=missing.types)

```


Exclude variable names for which NA>95%.  We have NAs when new_window is "no".  For feature extraction, the authors used a sliding window approach with different lengths from 0.5 second to 2.5 seconds, with 0.5 second overlap.
 
 
 
 
```{r explorationNA}
 
#searching for NA data and eliminating variables that have >97% NA/missing

Obstrain <- dim(training)[1]
Obstest <- dim(test)[1]

isNAtrain <- is.na(training)
isNAPct <- colSums(isNAtrain)/Obstrain
sum(isNAPct>.95)
sum(isNAPct==0)
isNAPctHIGH <-isNAPct[isNAPct>.97]
HIGHNAVAR <- names(isNAPctHIGH)
drops <- c(HIGHNAVAR)

training <- training[,!(names(training) %in% drops)]

```


``` {r nearzero search}
#searching for and eliminating variables used only for time stamp and record numbers.

library(caret)
library(randomForest)

nsv<-nearZeroVar(training,saveMetrics=TRUE) # identify variables that have little to no variability
sum(nsv$nzv==TRUE) #there is 1. It is new_window.  Also drop time stamp and record number related variables and num_window
drops2 <- c("X","raw_timestamp_part_1","raw_timestamp_part_2","cvtd_timestamp","new_window","num_window")
training <- training[,!(names(training) %in% drops2)]

```

```{r can we eliminate user name?}
#Test the hypothesis whether the class is independent of username at .05 significance level.
#H0: Variable A and Variable B are independent. 
#Ha: Variable A and Variable B are not independent.

#If the sample findings are unlikely, given the null hypothesis, the researcher rejects the null hypothesis. Typically, this involves comparing the P-value to the significance level, and rejecting the null hypothesis when the P-value is less than the significance level.


D <- with(training,table(user_name,classe))
chisq.test(D)

#As the p-value is extremely tiny (smaller than the .05 significance level), we reject the null hypothesis that the username is independent of the classe.  therefore we must keep it in.

```


``` {r explore high correlation variables}

M <- abs(cor(training[2:53]))
diag(M) <- 0 #change the 100% correlation to 0 so you can exclude variables correlated with themselves
highcorvar <- which(M > 0.9,arr.ind=TRUE) # which pairs have high correlation

#observe that 22 pairs have very high correlation and that its too difficult to pick which ones to eliminate and it wouldn't be that many to eliminate anyway.



```

```{r settingcrossvalidation}

set.seed(12345)
inTrain <- createDataPartition(y=training$classe,
                               p=0.7, list=FALSE)  
training2 <- training[inTrain,]
validation <- training[-inTrain,]
dim(training2); dim(validation)


```



```{r randomforest, cache = TRUE}

# for computational time saving, partition into a smaller subset

set.seed(23456)
inTrainRF <- createDataPartition(y=training2$classe,
                               p=0.25, list=FALSE)  
trainingRF <- training2[inTrainRF,]
testingRF <- training2[-inTrainRF,]

modFit <- train(classe ~ ., data = trainingRF, method="rf")

modFit$finalModel

#Call:
 #randomForest(x = x, y = y, mtry = param$mtry) 
 #              Type of random forest: classification
 #                    Number of trees: 500
#No. of variables tried at each split: 29

 #       OOB estimate of  error rate: 2.79%

predRF<-predict(modFit,testingRF)
confusionMatrix(predRF,testingRF$classe)

#Confusion Matrix and Statistics

 #         Reference
#Prediction    A    B    C    D    E
 #        A 2910   80    0    0    3
  #       B   14 1865   53    2    5
   #      C    3   45 1723   52   13
   #      D    1    3   21 1631   18
   #      E    1    0    0    4 1854

#Overall Statistics
                                          
 #              Accuracy : 0.9691          
 #                95% CI : (0.9656, 0.9724)
 #   No Information Rate : 0.2843          
 #   P-Value [Acc > NIR] : < 2e-16         
                                          
 #                 Kappa : 0.9609          
 #Mcnemar's Test P-Value : 1.9e-15    
```


```{r crossvalidation}

predRFCV<-predict(modFit,validation)
confusionMatrix(predRFCV,validation$classe)


# Accuracy : 0.967     



#ntree  Number of trees to grow. This should not be set to too small a number, to ensure that every input row gets predicted at least a few times. default is 500

#nodesize  Minimum size of terminal nodes. Setting this number larger causes smaller trees to be grown (and thus take less time). Note that the default values are different for classification (1) and regression (5). 

#mtry   Number of variables randomly sampled as candidates at each split. Note that the default values are different for classification (sqrt(p) where p is number of variables in x) and regression (p/3)

#importance=TRUE importance of predictors will be assessed




```

```{r for quiz}
#preprocess test; remove unneeded variables

testF <- test[,!(names(test) %in% drops)]
testF <- testF[,!(names(testF) %in% drops2)]

predQuiz<-predict(modFit,testF)

 #predQuiz
 #[1] B A B A A E D B A A B C B A E E A B B B
```

